Follow-up Samples

### 2017-04-10 submission/manuscript.Rmd

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
. includes:
. . in_header: header.tex
. keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: reference.bib
---

@@

## Abstract [ edit at end ]

@@

**Results.** There were large changes the the bacterial communities associated with treatment across the three groups. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r change_summary["thetayc", "Pvalue_adnVcrc"]`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value>0.05). Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value=`r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups. The OTUs that were shared across the models for the three diagnosis groups were dominated by lineages that are commonly considered to be commensals (e.g. *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).

4. Models suggest that the positive predictive value from diagnostic models decrease following treatment of carcinomas

Finally, we used a larger cohort that contained individuals with

Using diagnostic Random Forest models that were trained to differentiate between individuals with normal colons or those with adenomas, advanced adenomas, or carcinomas we measured the change in the positive probability of having one of the three diagnoses.

Using Random Forest models to assess whether changes in post-treatment samples were towards a normal community, only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma.

5. CRC-associated bacterial populations are unique and they decline in relative abundance with treatment

6. Among the patients with carcinoma, the type of treatment did not significantly alter the amount of change to their bacterial communities.

Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** Our data partially supports the hypothesis that the microbiome changes after treatment towards a normal community. Individuals with carcinoma had more drastic differences to the overall community than those with adenoma. Common OTUs to all models were overwhelmingly from commensal bacteria, suggesting that these bacteria may be important in initial polyp formation, development of advanced adenoma, and transition to carcinoma.

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009]. The rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010]. Despite this improvement, there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009]. Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014]. Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.

[ 	fix: the following paragraph is mixing two things - mechanistic studies and biomarker studies. they need to be split up so that you establish that there is a mechanistic link and that we can use biomarkers to diagnose people with adenomas and carcinomas from those with normal colons. ]

Recent studies in humans and mouse models have demonstrated a connection between the  that altered membership and structure of the gut microbiome correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014; @zackular_gut_2013]. Further, bacterial communities on the mucosa are altered between normal and tumor tissue [@dejea_microbiota_2014].
Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest the possibility of using biomarkers to diagnose disease.

Indeed, models created with microbiome data and clinical tests, such as Fecal Immunoglobulin Test (FIT), result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014]. Although these studies show how changed microbial communities or invasion by inflammatory species can impact CRC progression [@flynn_metabolic_2016], they provide very little information as to whether treatment to remove lesions affects the presence of bacterial populations associated with those lesions.

Providing an answer to this question is important because it has far reaching implications on both how the bacterial community causes the formation of more polyps [@zackular_manipulation_2016; @arthur_microbial_2014] and the ability to use the microbiome as a predictive screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. Understanding polyp formation and transition to advanced adenoma and then carcinoma is crucial to being able to understand how to prevent CRC occurrence. Response of the community to treatment is also equally important to predictive models designed for screening purposes since an unresponsive community would provide little additional information for important events, such as recurrence [@hassan_efficacy_2016].

Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes towards a normal microbiome in those with adenoma, advanced adenoma, and carcinoma. First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics. Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between pre- and post-treatment samples. We also used these models to assess whether changes in the community were toward a more normal microbiome. Finally, we assessed both whether surgery for adenomas and advanced adenomas provided larger community changes and whether chemotherapy or radiation provided additional changes to the microbiome over surgical resection. This study aims to provide evidence as to whether the altered CRC microbiome persists or shifts back towards normal after such interventions.

>>>>

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
  includes:
    in_header: header.tex
  keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: reference.bib
---

@@

## Abstract

@@

**Results.** There were large changes the the bacterial communities associated with treatment across the three groups. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r change_summary["thetayc", "Pvalue_adnVcrc"]`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value>0.05). Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value=`r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups. However, across the three models, there was little overlap between the OTUs that were indicative of treatment. Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent a treatment were more likely to have OTUs associated with normal colons. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses. Although we could clearly differentiate pre- and post-treatment communities from the three diagnosis groups, only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a carcinoma (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with chemotherapy and those treated with chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiome will yield improvements in disease management.

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009]. Disease mortality has significantly decreased, thanks to improvements in screening [@jemal_cancer_2010]. Despite this improvement, there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009]. Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014]. Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.

There is growing evidence that the gut microbiota is involved in the progression of CRC. Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [sears-ros/etbf into min mice, garrett-into min mice, han? tissue culture, @arthur_microbial_2014]. Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; ; @zackular_gut_2013; baxter human feces into mice]. Collectively, these studies hypothesize that the microbiota is altering the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].

Building on this evidence, several human-based studies have identified unique signatures of colonic lesions. One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014, garrett, jobin - science]. Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; zackular ]. These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions. They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions. Gut bacteria have a role in tumorigenesis and there is promise that these populations may be useful biomarkers for aiding in the early detection of disease.

Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota. If the community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence. To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner? If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?

We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre- and post-treatment. We used both classical community analysis to compare the alpha and beta-diversity of communities pre- and post treatment. Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group. Finally, we measured the predictive probability of diagnosis models to determine whether treatment would alter the probability of being diagnosed as having a normal colon or the colon of the original diagnosis group. Ultimately, we found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas. Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence.
